---
title: "Basic choropleth"
author: "Hugo Janssen (nl-hugo)"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: html_document
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  dpi = 300,
  fig.width = 6,
  fig.height = 5,
  fig.path = "img/basic-",
  strip.white = T,
  dev = "png",
  dev.args = list(png = list(bg = "transparent")),
  echo = TRUE, warning = FALSE, message = FALSE
)
```


This document creates a basic choropleth from a CBS shapefile containing data about The Netherlands. Data provided by the Dutch Statistics Center [CBS](https://www.cbs.nl/en-gb) and [Kadaster](https://www.kadaster.com/). 


### Setup

Load the required packages and specify the file locations.

```{r}
#install.packages(c("rgeos", "maptools", "ggplot2", "plyr", "ggthemes"))
library(rgeos)
library(maptools)
library(ggplot2)
library(ggthemes)
library(plyr)

# specify file locations
URL <-"https://www.cbs.nl/-/media/_pdf/2017/36/buurt_2015.zip"
datadir <- "data"

```


### Download

Download the zipped data file and unzip. Note that files are downloaded only when no data folder is present. 

```{r}
# create a dir and download if no datadir is present 
# assumes that files are downloaded if a datadir is present
if (!file.exists(datadir)) {
  dir.create(datadir)

  tmpfile <- paste(datadir, basename(URL), sep = "/")
  download.file(URL, destfile = tmpfile)
  unzip(tmpfile, exdir="./data")
}
```


### Prepare data

Read the topology from the shapefile and turn it into a dataframe that can be plotted by `ggplot`. Note that areas that are marked as sea or lake (by the `WATER` property) are excluded, so that coast is displayed nicely.

```{r}
# prepare filenames
basename <- gsub("buurt", "gem", tools::file_path_sans_ext(basename(URL)))
shpfile <- paste(basename, "shp", sep = ".")

# create spatial object
nl <- maptools::readShapeSpatial(paste(datadir, shpfile, sep = "/"))

# remove watery areas
nl <- nl[nl@data$WATER == "NEE", ]

# create id to join data
nl@data$id = as.character(nl@data$GM_CODE)

# convert to ggplot object
nl.points <- fortify(nl, region = "GM_CODE")

# join the geo points and the data
nl.df <- join(nl.points, nl@data, by = c("GM_CODE" = "id"))
```



### Plot

Create a basic plot of the map with minimal styling.

```{r choropleth}
# create the plot
p <- ggplot(nl.df) + 
  aes(x = long, y = lat, group = group, fill = P_GEBOO) + 
  geom_polygon() +
  geom_path(colour = "white", size = 0.25) +
  scale_fill_distiller(palette = "PiYG") + 
  labs(x = NULL, 
       y = NULL, 
       title = "Geboortecijfer per gemeente",
       subtitle = "Aantal geboortes per 1000 inwoners") +
  coord_equal() +
  theme_minimal() +
  theme (
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    axis.text = element_blank()
  )

p
```

